 

### 2020-12-31 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@nakamuraForestsTheirCanopies2017; @ozanneBiodiversityMeetsAtmosphere2003] by buffering extreme temperatures along with other macroclimatic conditions in the understory. This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient.[Bertrand et al. 2011]

The capacity of buffering largly depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions.
Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018;@huangAirTemperatureOptima2019]. Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests [ref]. Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a]. Evidence shows that larger trees suffer during drought resulting in heat stressed canopy-leaves, which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to occur [@bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a; @mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continues evapotranspiration and consequential buffering of the understory [REF]

While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming. As a result, understories experience a different set of microclimatic pressures, e.g limited light and related foliar traits, inter-species competition etc.. with rising temperature [@bartholomewSmallTropicalForest2020]. Since temperature is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011]. Evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013]. Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [@niinemetsControversyTraitsConferring2006] which may have further impacts on understory species with rising temperatures. 
Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellPervasiveShiftsForest2020].
Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020]. Each of these changes have potential feedbacks to climate change.

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests during wet season tended to have higher $T_{air}$, and $VPD$ in upper canopy and canopy gaps than lower in understory where $T_{air}$ and $VPD$ were lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

>>>>

Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. This creates a vertical stratification of biophysical environments such as temperature, light, wind, humidity and Co2 which influences leaf temperature and metabolism along the gradient.[Bertrand et al. 2011]

The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions.
Increasing temperature has profound impacts on canopy cover such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Additionally, increased temperature is driving more frequent droughts causing hydraulic limitation in forests [ref]. Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water and precipitation [@rollinsonClimateSensitivityUnderstory 2020]. Evidence shows that larger trees suffer during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to occur [@@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continues evapotranspiration and consequential buffering of the understory [REF]

While forest microclimates are insulated from the extremes of macroclimate warming, they are still experiencing warming [@bertrandCommentForestMicroclimate2020]. As a result, understories experience a different set of microclimatic pressures with rising temperature in combination with conditions of limited light and related foliar traits, inter-species resource competition etc.. [@bartholomewSmallTropicalForest2020]. Since temperature sensitivity is directly linked with plant metabolic processes and growth, even around 1 degree of warming can lead to changes in plant species composition and growth [Bertrand et al.2011]. Tree-ring evidence shows reduced growth of understory trees relative to overstory with warming temperatures, in mesic temperate forests[@rollinsonClimateSensitivityUnderstory, 2020], along with other studies that show thermophilization in plant communities and reshuffling of species acclimated to cooler [mean annual] temperatures[@duqueThermophilizationAdultJuvenile2015; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013; @suggittExtinctionRiskClimate2018; @scheffersMicrohabitatsReduceAnimal2013]. Considering more frequent occurences of drought, extensive analysis also shows negetive correlation between species shade tolerance and drought tolerance globally [@niinemetsControversyTraitsConferring2006] which may have further impacts on understory species with rising temperatures. 
Microclimate warming has implications for forest regeneration[@vonarxSpatiotemporalEffectsForest2012a] as temperature driven photosynthesis limitation is leading to reduced tree growth rates, driving a global trend towards disequillibrium in forests with younger trees [@mcdowellPervasiveShiftsForest2020].
Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020]. Each of these changes have potential feedbacks to climate change.

@@

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests during wet season tended to have higher $T_{air}$, and $VPD$ in upper canopy and canopy gaps than lower in understory where $T_{air}$ and $VPD$ were lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]. In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas as understory was insensitive [[@rollinsonClimateSensitivityUnderstory, 2020]
